Modelling at-bats in baseball using the generalised Pareto distribution

### 2023-02-20 bibat/examples/baseball/docs/report.qmd
Update case study

I thought I'd test this out with some sports data, and luckily there is a really
nice baseball example on [the Stan case studies
website](https://mc-stan.org/users/documentation/case-studies/pool-binary-trials.html),
complete with [data from the 2006 Major league
season])https://github.com/stan-dev/example-models/blob/master/knitr/pool-binary-trials/baseball-hits-2006.csv).
After I posted some early results [on the Stan discourse
forum](https://discourse.mc-stan.org/t/baseball-analysis-using-latent-generalised-pareto-distribution/29222),
user jd_c kindly gathered and posted [even more
data](https://discourse.mc-stan.org/uploads/short-url/cRJtia2jKxiX01ZEraHUVK7jbHI.RData),
covering 5 full major league baseball seasons.

@@

I added a line with the text `pyreadr` to the file `requirements.txt`, then
added the following code to a new script called `fetch_data.py`:

@@

@@

    player: pa.typing.Series[str]
    season: pa.typing.Series[str]
    n_attempt: pa.typing.Series[int] = pa.Field(ge = 1)
    n_success: pa.typing.Series[int] = pa.Field(ge = 0)
```

Next we can write some functions that create `PreparedData` objects. These
should live in the file `data_preparation_functions.py`. In this case we write a
couple of data preparation functions: `prepare_data_2006` and
`prepare_data_recent`:

@@

To check that all this works, we can run the script `prepare_data.py` manually
or using `make analysis`. Now if we look in the file
`data/prepared/recent/measurements.csv` we should see some lines that look like
this:

```{csv}
,player,season,n_attempt,n_success
0,abreujo02,2018,553,180
1,acunaro01,2018,487,178
3,adamewi01,2018,322,112
6,adamsla01,2018,29,10
7,adamsma01,2018,337,104
8,adamsma01,2018,277,92
9,adamsma01,2018,60,12
```

@@

# Extending the analysis to the recent data

To model the more recent data, all I had to do was create some new inference
folders with appropriate `prepared_data_dir` fields in their `config.toml`
files. For example, here is the `config.toml` file for the `gparetorecent`
inference:

```{.toml include="../inferences/gparetorecent/config.toml"}
```

@@

I think this is very interesting. Both models' prior distributions had similar
regularisation levels, and they more or less agree about the abilities of the
players with the most at-bats, both in terms of locations and the widths of the
plausible intervals for the true success rate. However, the models ended up with
dramatically different certainty levels about the abilities of players with few
at-bats. This pattern was true both for the small 2006 dataset and the much
larger 'recent' dataset.

>>>>

I thought I'd test this out with some sports data, and luckily there is a really
nice baseball example on [the Stan case studies
website](https://mc-stan.org/users/documentation/case-studies/pool-binary-trials.html),
complete with [data from the 2006 Major league
season])https://github.com/stan-dev/example-models/blob/master/knitr/pool-binary-trials/baseball-hits-2006.csv).
After I posted some early results [on the Stan discourse
forum](https://discourse.mc-stan.org/t/baseball-analysis-using-latent-generalised-pareto-distribution/29222),
other users suggested that it might be interesting to model similar data from
different seasons. This data can now be found quite easily on the [baseball
databank](https://baseballdb.lawlesst.net/).

@@

To fetch raw data from the internet, I wrote a new script called `fetch_data.py`:

@@

Since this worked, I added a new makefile target for the raw data files:

```{Makefile}
RAW_DATA = data/raw/2006.csv data/raw/bdb-main.csv data/raw/bdb-post.csv data/raw/bdb-apps.csv
...
$(RAW_DATA): $(ACTIVATE_VENV) && python fetch_data.py
```

@@

    player_season: pa.typing.Series[str]
    season: pa.typing.Series[str]
    n_attempt: pa.typing.Series[int] = pa.Field(ge = 1)
    n_success: pa.typing.Series[int] = pa.Field(ge = 0)
```

Next we can write some functions that create `PreparedData` objects. These
should live in the file `data_preparation_functions.py`. In this case we write a
couple of data preparation functions: `prepare_data_2006` and
`prepare_data_bdb`:

@@

To check that all this works, we can run the script `prepare_data.py` manually
or using `make analysis`. Now if we look in the file
`data/prepared/bdb/measurements.csv` we should see some lines that look like
this:

```{csv}
,player_season,season,n_attempt,n_success
0,abreujo02,2018,553,180
1,acunaro01,2018,487,178
3,adamewi01,2018,322,112
6,adamsla01,2018,29,10
7,adamsma01,2018,337,104
8,adamsma01,2018,277,92
9,adamsma01,2018,60,12
```

@@

# Extending the analysis to the baseballdatabank data

To model the more recent data, all I had to do was create some new inference
folders with appropriate `prepared_data_dir` fields in their `config.toml`
files. For example, here is the `config.toml` file for the `gparetobdb`
inference:

```{.toml include="../inferences/gparetobdb/config.toml"}
```

@@

I think this is very interesting. Both models' prior distributions had similar
regularisation levels, and they more or less agree about the abilities of the
players with the most at-bats, both in terms of locations and the widths of the
plausible intervals for the true success rate. However, the models ended up with
dramatically different certainty levels about the abilities of players with few
at-bats. This pattern was true both for the small 2006 dataset and the much
larger baseballdatabank dataset.
